---
title: "Stat 184 4.3 QMD Andrew McRavey"
Date: "4/11/2026"
author: "Andrew McRavey"
format: pdf
editor: visual
---

# Part 1: Busiest Airports Analysis

## 1.1 Tidying/Table Creation Function:

```{r}

library(tidyverse)

airports_wide <- tribble(
  ~iata_code, ~airport_name,                      ~city,              ~country,         ~region,
  ~`2020`, ~`2021`, ~`2022`, ~`2023`, ~`2024`, ~`2025`,
  
  "ATL", "Hartsfield-Jackson Atlanta Intl", "Atlanta",           "United States",  "North America",
  42.92,   75.70,   93.70,  104.65,  108.07,  106.30,
  
  "FRA", "Frankfurt Airport",               "Frankfurt",         "Germany",        "Europe",
  18.77,   24.81,   48.92,   59.36,   61.56,   63.19,
  
  "PKX", "Beijing Daxing Intl",             "Beijing",           "China",          "East Asia",
  16.00,   25.00,   10.23,   39.41,   49.42,   56.00,
  
  "DXB", "Dubai International",             "Dubai",             "UAE",            "Middle East",
  25.84,   29.11,   66.10,   86.90,   92.30,   95.20,
  
  "LHR", "London Heathrow",                 "London",            "United Kingdom", "Europe",
  22.11,   19.39,   61.60,   79.20,   83.86,   84.46,
  
  "DFW", "Dallas/Fort Worth Intl",          "Dallas-Fort Worth", "United States",  "North America",
  39.40,   62.50,   73.40,   81.80,   87.80,   90.00
)


# 2. Pivot to tidy long format
#    One row = one airport × one year

airports_tidy <- airports_wide |>
  pivot_longer(
    cols      = `2020`:`2025`,
    names_to  = "year",
    values_to = "passengers_millions"
  ) |>
  mutate(year = as.integer(year)) |>
  arrange(iata_code, year)


# 3. Add year-over-year change columns (within each airport)

airports_tidy <- airports_tidy |>
  group_by(iata_code) |>
  mutate(
    yoy_change_millions = passengers_millions - lag(passengers_millions),
    yoy_pct_change      = round((passengers_millions / lag(passengers_millions) - 1) * 100, 1)
  ) |>
  ungroup()

# 4. Enforce factor ordering for airports (by 2024 traffic)

airport_order <- airports_tidy |>
  filter(year == 2024) |>
  arrange(desc(passengers_millions)) |>
  pull(iata_code)

airports_tidy <- airports_tidy |>
  mutate(iata_code = factor(iata_code, levels = airport_order))


# 5. Preview

glimpse(airports_tidy)
print(airports_tidy, n = 36)
```

## 1.2 Plotting Function:

```{r}

library(tidyverse)

# Plot-specific wrangling: label data for direct line labels

label_data <- airports_tidy |>
  filter(year == 2025)


# Plot

ggplot(airports_tidy, aes(x = year, y = passengers_millions,
                          color = iata_code, group = iata_code)) +
  
  # COVID dip shading
  annotate("rect",
           xmin = 2019.5, xmax = 2021.5,
           ymin = -Inf,   ymax = Inf,
           fill = "grey90", alpha = 0.5) +
  annotate("text",
           x = 2020.5, y = 108,
           label = "COVID-19\nrestrictions",
           size = 3, color = "grey50", fontface = "italic") +
  
  # Lines and points
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  
  # Direct endpoint labels (replaces legend)
  geom_text(data  = label_data,
            aes(label = paste0(iata_code, "\n", passengers_millions, "M")),
            hjust  = -0.1,
            size   = 3.2,
            fontface = "bold",
            show.legend = FALSE) +
  
  # Axis formatting
  scale_x_continuous(
    breaks = 2020:2025,
    expand = expansion(mult = c(0.03, 0.16))   # right margin for labels
  ) +
  scale_y_continuous(
    breaks = seq(0, 120, by = 20),
    labels = scales::label_number(suffix = "M"),
    limits = c(0, 120)
  ) +
  scale_color_brewer(palette = "Dark2") +
  
  # Labels
  labs(
    title    = "Annual Passenger Traffic at Six Major Airports (2020–2025)",
    subtitle = "Total passengers (enplaned + deplaned + direct transit), in millions",
    x        = "Year",
    y        = "Passengers (millions)",
    caption  = "Source: Wikipedia — List of Busiest Airports by Passenger Traffic\n2025 figures are preliminary estimates published by each airport's operator."
  ) +
  
  # Theme
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey40", margin = margin(b = 10)),
    plot.caption     = element_text(color = "grey50", size = 8, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line.x      = element_line(color = "grey70"),
    legend.position  = "none"
  )

```

## 1.3 Busiest Airports Narrative:

This graph help effectively show the effects of COVID-19 restrictions on airports throughout the world. As you can see on the left sied of the graph, where it is shaded, almost all airpots have the least amount of traffic due to the increased regulations on travel due to COVID-19. as time progresses and regulations are lifted air traffic returns to normal. One aspect of this graph that is interesting in the amount of passengers at Beijing (BKX) being lowest in 2022, This is because of China's stricter COVID policy and the fact that many regulations remained in China during 2022 which made BKX suffer some consequences because of this. However it was able to recover to a similar level to other airports on this list by 2025. All in all this plot is a powerful resource that shows the effect of long term global pandemics and policy on airport traffic.

# Part 2: Monte Carlo Numerical Integration

## 2.1 Monte Carlo Visualizations Function:

```{r}

monte_carlo_sim <- function(n, x_min, x_max, y_min, y_max, f) {
  x_vals <- runif(n, min = x_min, max = x_max)
  y_vals <- runif(n, min = y_min, max = y_max)
  
  df <- data.frame(
    x = x_vals,
    y = y_vals
  )
  
  df$fx <- f(df$x)
  df$inside <- ifelse(df$y <= df$fx, TRUE, FALSE)
  
  rect_area <- (x_max - x_min) * (y_max - y_min)
  proportion_inside <- mean(df$inside)
  estimate <- proportion_inside * rect_area
  
  attr(df, "estimate") <- estimate
  return(df)
}

library(ggplot2)
library(dplyr)

f <- function(x) dbeta(x, shape1 = 2, shape2 = 5)

df <- monte_carlo_sim(
  n = 1000,
  x_min = 0,
  x_max = 1,
  y_min = 0,
  y_max = 1.5,   
  f = f
)

df <- df %>%
  mutate(category = ifelse(inside, "Below Curve", "Above Curve"))

estimate <- attr(df, "estimate")

curve_data <- data.frame(
  x = seq(0, 1, length.out = 200)
) %>%
  mutate(y = f(x))

plot1 <- ggplot() +
  geom_point(data = df, aes(x = x, y = y, color = category), alpha = 0.6) +
  geom_line(data = curve_data, aes(x = x, y = y), color = "black", size = 1) +
  labs(
    title = paste("Monte Carlo Integration (n = 1000)"),
    subtitle = paste("Estimated Area =", round(estimate, 4)),
    x = "x",
    y = "y",
    color = "Point Type"
  ) +
  theme_minimal()

print(plot1)
```

## 2.2 Monte Carlo Simulation Narrative:

From my graph you can learn the exact value integral for the beta distribution or at least something pretty close. The number I got in my highest sample size plots was around .8 and due to the law of large numbers I think this is around the correct result assuming my code is correct. This is a powerful tool to estimate definite integrals that may be analytically intractable or tedious to compute by hand. Rather than solving the integral symbolically, Monte Carlo integration leverages randomness uniformly scattering points across a bounding rectangle and computing the fraction that fall below the curve.

# Part 3: Planning and Prompting GenAI Tools

I will create a function to create tidy table and plot using Deepseek using data from the Calcium.csv file outlined in assignment 4.3. For this data set i will split it into 2 groups, A and B, separated by how they appear in the csv file, 2 groups of y0-y3, and plot them by arbitrary time values, 0-3 to make sense of the numbers in this undefined data set. I will make the x axis this arbitrary time value. I will make the y axis as a measurement value to make sense of each value under each column.

## 3.1 Prompt and Result

The prompt I used was: create code in r to make this attached file into a graph using each value as a measurement value, make it into 2 separate groups and compare the 2 on the graph. make sure to make a tidy table to make the code more understandable.

```{r}
# Load required libraries
library(tidyverse)
library(gridExtra)

# ------------------------------------------------------------------------------
# STEP 1: READ THE DATA
# ------------------------------------------------------------------------------

# Read the CSV file
calcium_data <- read.csv('C:/Users/andre/Downloads/calcium.csv', check.names = FALSE)

# View first few rows to understand structure
head(calcium_data)

# ------------------------------------------------------------------------------
# STEP 2: TIDY THE DATA (Convert from wide to long format)
# ------------------------------------------------------------------------------

# The data has 8 columns: 
# First 4 columns (y0, y1, y2, y3) = Group 1
# Last 4 columns (y0, y1, y2, y3) = Group 2

# Extract Group 1 columns (first 4 columns)
group1_data <- calcium_data[, 1:4]
names(group1_data) <- c("y0", "y1", "y2", "y3")

# Extract Group 2 columns (last 4 columns)
group2_data <- calcium_data[, 5:8]
names(group2_data) <- c("y0", "y1", "y2", "y3")

# Add participant ID to each group
group1_data$participant_id <- 1:nrow(group1_data)
group2_data$participant_id <- 1:nrow(group2_data)

# Convert Group 1 to long format
group1_long <- group1_data %>%
  pivot_longer(
    cols = starts_with("y"),
    names_to = "time_point",
    values_to = "measurement"
  ) %>%
  mutate(
    group = "Group A",
    time_numeric = as.numeric(gsub("y", "", time_point))
  ) %>%
  filter(!is.na(measurement))  # Remove any NA values

# Convert Group 2 to long format
group2_long <- group2_data %>%
  pivot_longer(
    cols = starts_with("y"),
    names_to = "time_point",
    values_to = "measurement"
  ) %>%
  mutate(
    group = "Group B",
    time_numeric = as.numeric(gsub("y", "", time_point))
  ) %>%
  filter(!is.na(measurement))  # Remove any NA values

# Combine both groups into one tidy table
tidy_calcium_data <- bind_rows(group1_long, group2_long)

# Display the tidy table structure
print("=== TIDY DATA TABLE ===")
print(head(tidy_calcium_data, 10))
print("\n")
print("Table structure:")
str(tidy_calcium_data)

# ------------------------------------------------------------------------------
# STEP 3: CALCULATE SUMMARY STATISTICS
# ------------------------------------------------------------------------------

# Calculate mean and standard deviation for each group at each time point
summary_stats <- tidy_calcium_data %>%
  group_by(group, time_point, time_numeric) %>%
  summarise(
    mean_value = mean(measurement, na.rm = TRUE),
    sd_value = sd(measurement, na.rm = TRUE),
    n = n(),
    se_value = sd_value / sqrt(n),  # Standard error
    .groups = "drop"
  )

print("\n=== SUMMARY STATISTICS ===")
print(summary_stats)

# ------------------------------------------------------------------------------
# STEP 4: CREATE COMPARISON GRAPHS
# ------------------------------------------------------------------------------

# Graph 1: Boxplot comparison at each time point
graph_boxplot <- ggplot(tidy_calcium_data, aes(x = time_point, y = measurement, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1.5) +
  stat_summary(fun = mean, geom = "point", position = position_dodge(width = 0.75), 
               shape = 23, size = 3, fill = "white") +
  labs(
    title = "Comparison of Calcium Measurements: Group A vs Group B",
    subtitle = "Boxplots show distribution, white diamonds show means",
    x = "Time Point",
    y = "Measurement Value",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral"))

# Graph 2: Line plot showing individual trajectories + group mean
graph_individual <- ggplot(tidy_calcium_data, aes(x = time_numeric, y = measurement, group = participant_id)) +
  geom_line(aes(color = group), alpha = 0.3, size = 0.5) +
  geom_point(aes(color = group), alpha = 0.3, size = 1) +
  stat_summary(aes(color = group), fun = mean, geom = "line", size = 1.5) +
  stat_summary(aes(color = group), fun = mean, geom = "point", size = 3) +
  labs(
    title = "Individual Trajectories with Group Means",
    subtitle = "Faint lines = individual participants, Thick lines = group averages",
    x = "Time Point",
    y = "Measurement Value",
    color = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_x_continuous(breaks = 0:3, labels = c("y0", "y1", "y2", "y3")) +
  scale_color_manual(values = c("Group A" = "blue", "Group B" = "red"))

# Graph 3: Bar chart with error bars (mean ± SE)
graph_bars <- ggplot(summary_stats, aes(x = time_point, y = mean_value, fill = group)) +
  geom_bar(stat = "identity", position = position_dodge(0.9), alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_value - se_value, ymax = mean_value + se_value),
                position = position_dodge(0.9), width = 0.25) +
  labs(
    title = "Mean Calcium Measurements (± Standard Error)",
    subtitle = "Bars show means, error bars show ±1 SE",
    x = "Time Point",
    y = "Mean Measurement Value",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral"))

# Graph 4: Violin plot with jittered points (shows distribution well)
graph_violin <- ggplot(tidy_calcium_data, aes(x = time_point, y = measurement, fill = group)) +
  geom_violin(alpha = 0.5, position = position_dodge(0.9)) +
  geom_jitter(aes(color = group), position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9),
              alpha = 0.4, size = 1) +
  stat_summary(fun = mean, geom = "point", position = position_dodge(0.9),
               shape = 18, size = 4, color = "black") +
  labs(
    title = "Distribution of Calcium Measurements",
    subtitle = "Violin plots show density, points show individual values, black diamonds show means",
    x = "Time Point",
    y = "Measurement Value",
    fill = "Group",
    color = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral")) +
  scale_color_manual(values = c("Group A" = "blue", "Group B" = "red"))

# ------------------------------------------------------------------------------
# STEP 5: DISPLAY ALL GRAPHS
# ------------------------------------------------------------------------------

# Print each graph individually
print(graph_boxplot)
print(graph_individual)
print(graph_bars)
print(graph_violin)

# ------------------------------------------------------------------------------
# STEP 6: STATISTICAL COMPARISON
# ------------------------------------------------------------------------------

# Perform t-tests comparing groups at each time point
stat_comparisons <- tidy_calcium_data %>%
  group_by(time_point) %>%
  summarise(
    group_A_mean = mean(measurement[group == "Group A"]),
    group_B_mean = mean(measurement[group == "Group B"]),
    difference = group_A_mean - group_B_mean,
    t_statistic = t.test(measurement ~ group)$statistic,
    p_value = t.test(measurement ~ group)$p.value,
    significance = ifelse(p_value < 0.05, "Significant", "Not significant")
  )

print("\n=== STATISTICAL COMPARISONS (t-tests) ===")
print(stat_comparisons)

```

## 3.2 Planning and Prompting GenAI Tools Narrative:

This code helps to understand this admittedly barren data set and create some context behind the data. As for what you understand form this data set, There is a noticeable decline in the measurements for both groups around y3 suggesting some natural phenomenon happens around y3 that leads to smaller meaurements and less variance.

# Part 4: Reflection

## 4.1 What I have learned in the course:

In this course I have learned how to wrangle data to create tidy tables and how create effective data visualizations using R. Also I have learned the importance of documentation in coding.

One example of my culmination of learning is in the monte carlo simulation code. I think I did a good job in creating effective code to make an understandable and appealing looking visualization that is easily scalable to high n values. I would not be able to code this effectively in R without the things I learned in this course and I am proud of how it turned out.

Another example of my learning from this course is this QMD file, I was unfamiliar with Quarto before this assignment and class. I have learned how important documentation is in coding because of this class and this assignment. From this assignment I have learned how to make coding easier to understand and follow using nicely formatted and digestible sections.

Finally this class was important for me to learn how to use GenAI effectively, in both the assignment where we used AI to create alt text and long descriptions and for this assignment where AI use is much more in depth in the actual coding process. This class showed hwo useful AI is as a tool for coding, tiding, summarizing, and plotting data.

# Part 5: Appendixes

## 5.1 AI use Appendix

I used GenAI in part 4 of this assignment as directed by the instructor to put my plan for an original data visualization into effect.

GenAI usage information:

Model Used: Deepseek

Date Used: 4/13/2026

Exact Prompt and Response:

create code in r to make this attached file into a graph using each value as a measurement value, make it into 2 separate groups and compare the 2 on the graph. make sure to make a tidy table to make the code more understandble.

Here's complete R code to tidy your data and create a clear comparison graph between the two groups.

r

```         
# Load required libraries library(tidyverse) library(gridExtra)  # ------------------------------------------------------------------------------ # STEP 1: READ THE DATA # ------------------------------------------------------------------------------  # Read the CSV file calcium_data <- read.csv("calcium.csv", check.names = FALSE)  # View first few rows to understand structure head(calcium_data)  # ------------------------------------------------------------------------------ # STEP 2: TIDY THE DATA (Convert from wide to long format) # ------------------------------------------------------------------------------  # The data has 8 columns:  # First 4 columns (y0, y1, y2, y3) = Group 1 # Last 4 columns (y0, y1, y2, y3) = Group 2  # Extract Group 1 columns (first 4 columns) group1_data <- calcium_data[, 1:4] names(group1_data) <- c("y0", "y1", "y2", "y3")  # Extract Group 2 columns (last 4 columns) group2_data <- calcium_data[, 5:8] names(group2_data) <- c("y0", "y1", "y2", "y3")  # Add participant ID to each group group1_data$participant_id <- 1:nrow(group1_data) group2_data$participant_id <- 1:nrow(group2_data)  # Convert Group 1 to long format group1_long <- group1_data %>%   pivot_longer(     cols = starts_with("y"),     names_to = "time_point",     values_to = "measurement"   ) %>%   mutate(     group = "Group A",     time_numeric = as.numeric(gsub("y", "", time_point))   ) %>%   filter(!is.na(measurement))  # Remove any NA values  # Convert Group 2 to long format group2_long <- group2_data %>%   pivot_longer(     cols = starts_with("y"),     names_to = "time_point",     values_to = "measurement"   ) %>%   mutate(     group = "Group B",     time_numeric = as.numeric(gsub("y", "", time_point))   ) %>%   filter(!is.na(measurement))  # Remove any NA values  # Combine both groups into one tidy table tidy_calcium_data <- bind_rows(group1_long, group2_long)  # Display the tidy table structure print("=== TIDY DATA TABLE ===") print(head(tidy_calcium_data, 10)) print("\n") print("Table structure:") str(tidy_calcium_data)  # ------------------------------------------------------------------------------ # STEP 3: CALCULATE SUMMARY STATISTICS # ------------------------------------------------------------------------------  # Calculate mean and standard deviation for each group at each time point summary_stats <- tidy_calcium_data %>%   group_by(group, time_point, time_numeric) %>%   summarise(     mean_value = mean(measurement, na.rm = TRUE),     sd_value = sd(measurement, na.rm = TRUE),     n = n(),     se_value = sd_value / sqrt(n),  # Standard error     .groups = "drop"   )  print("\n=== SUMMARY STATISTICS ===") print(summary_stats)  # ------------------------------------------------------------------------------ # STEP 4: CREATE COMPARISON GRAPHS # ------------------------------------------------------------------------------  # Graph 1: Boxplot comparison at each time point graph_boxplot <- ggplot(tidy_calcium_data, aes(x = time_point, y = measurement, fill = group)) +   geom_boxplot(alpha = 0.7, outlier.size = 1.5) +   stat_summary(fun = mean, geom = "point", position = position_dodge(width = 0.75),                 shape = 23, size = 3, fill = "white") +   labs(     title = "Comparison of Calcium Measurements: Group A vs Group B",     subtitle = "Boxplots show distribution, white diamonds show means",     x = "Time Point",     y = "Measurement Value",     fill = "Group"   ) +   theme_minimal() +   theme(     plot.title = element_text(hjust = 0.5, face = "bold"),     plot.subtitle = element_text(hjust = 0.5),     legend.position = "bottom"   ) +   scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral"))  # Graph 2: Line plot showing individual trajectories + group mean graph_individual <- ggplot(tidy_calcium_data, aes(x = time_numeric, y = measurement, group = participant_id)) +   geom_line(aes(color = group), alpha = 0.3, size = 0.5) +   geom_point(aes(color = group), alpha = 0.3, size = 1) +   stat_summary(aes(color = group), fun = mean, geom = "line", size = 1.5) +   stat_summary(aes(color = group), fun = mean, geom = "point", size = 3) +   labs(     title = "Individual Trajectories with Group Means",     subtitle = "Faint lines = individual participants, Thick lines = group averages",     x = "Time Point",     y = "Measurement Value",     color = "Group"   ) +   theme_minimal() +   theme(     plot.title = element_text(hjust = 0.5, face = "bold"),     plot.subtitle = element_text(hjust = 0.5),     legend.position = "bottom"   ) +   scale_x_continuous(breaks = 0:3, labels = c("y0", "y1", "y2", "y3")) +   scale_color_manual(values = c("Group A" = "blue", "Group B" = "red"))  # Graph 3: Bar chart with error bars (mean ± SE) graph_bars <- ggplot(summary_stats, aes(x = time_point, y = mean_value, fill = group)) +   geom_bar(stat = "identity", position = position_dodge(0.9), alpha = 0.7) +   geom_errorbar(aes(ymin = mean_value - se_value, ymax = mean_value + se_value),                 position = position_dodge(0.9), width = 0.25) +   labs(     title = "Mean Calcium Measurements (± Standard Error)",     subtitle = "Bars show means, error bars show ±1 SE",     x = "Time Point",     y = "Mean Measurement Value",     fill = "Group"   ) +   theme_minimal() +   theme(     plot.title = element_text(hjust = 0.5, face = "bold"),     plot.subtitle = element_text(hjust = 0.5),     legend.position = "bottom"   ) +   scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral"))  # Graph 4: Violin plot with jittered points (shows distribution well) graph_violin <- ggplot(tidy_calcium_data, aes(x = time_point, y = measurement, fill = group)) +   geom_violin(alpha = 0.5, position = position_dodge(0.9)) +   geom_jitter(aes(color = group), position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9),               alpha = 0.4, size = 1) +   stat_summary(fun = mean, geom = "point", position = position_dodge(0.9),                shape = 18, size = 4, color = "black") +   labs(     title = "Distribution of Calcium Measurements",     subtitle = "Violin plots show density, points show individual values, black diamonds show means",     x = "Time Point",     y = "Measurement Value",     fill = "Group",     color = "Group"   ) +   theme_minimal() +   theme(     plot.title = element_text(hjust = 0.5, face = "bold"),     plot.subtitle = element_text(hjust = 0.5),     legend.position = "bottom"   ) +   scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral")) +   scale_color_manual(values = c("Group A" = "blue", "Group B" = "red"))  # ------------------------------------------------------------------------------ # STEP 5: DISPLAY ALL GRAPHS # ------------------------------------------------------------------------------  # Print each graph individually print(graph_boxplot) print(graph_individual) print(graph_bars) print(graph_violin)  # ------------------------------------------------------------------------------ # STEP 6: STATISTICAL COMPARISON # ------------------------------------------------------------------------------  # Perform t-tests comparing groups at each time point stat_comparisons <- tidy_calcium_data %>%   group_by(time_point) %>%   summarise(     group_A_mean = mean(measurement[group == "Group A"]),     group_B_mean = mean(measurement[group == "Group B"]),     difference = group_A_mean - group_B_mean,     t_statistic = t.test(measurement ~ group)$statistic,     p_value = t.test(measurement ~ group)$p.value,     significance = ifelse(p_value < 0.05, "Significant", "Not significant")   )  print("\n=== STATISTICAL COMPARISONS (t-tests) ===") print(stat_comparisons)  # ------------------------------------------------------------------------------ # STEP 7: SAVE THE BEST GRAPH (optional) # ------------------------------------------------------------------------------  # Save the boxplot as a high-resolution image ggsave("calcium_comparison_boxplot.png", graph_boxplot, width = 10, height = 6, dpi = 300)  # Save the tidy data for future use write.csv(tidy_calcium_data, "tidy_calcium_data.csv", row.names = FALSE)  # ------------------------------------------------------------------------------ # BONUS: Interactive visualization (if plotly is installed) # ------------------------------------------------------------------------------  # Uncomment this section if you have plotly installed # library(plotly) # interactive_plot <- ggplotly(graph_individual) # interactive_plot  print("\n=== ANALYSIS COMPLETE ===") print("Tidy data saved to 'tidy_calcium_data.csv'") print("Boxplot saved to 'calcium_comparison_boxplot.png'")
```

This code will:

1.  **Read your CSV file** and handle the structure correctly

2.  **Create a tidy table** with columns: participant_id, time_point, measurement, group, time_numeric

3.  **Generate 4 different graphs** comparing Group A vs Group B:

    -   Boxplot with mean points

    -   Individual trajectories with group means (line plot)

    -   Bar chart with error bars

    -   Violin plot with jittered points

4.  **Perform statistical comparisons** (t-tests) at each time point

5.  **Save** the tidy data and the best graph to your computer

The tidy data table makes the code easy to understand because each row represents a single measurement, making it straightforward to use with ggplot2. Choose the graph that best tells your data story - I recommend the **boxplot** or **individual trajectories** plot for clear group comparison.

## 5.2 Code Appendix

## 5.2.1 Busiest Airport Code

```{r}
library(tidyverse)

# ------------------------------------------------------------
# 1. Enter raw data directly from Wikipedia's table
#    (wide format, one column per year)
# ------------------------------------------------------------
airports_wide <- tribble(
  ~iata_code, ~airport_name,                      ~city,              ~country,         ~region,
  ~`2020`, ~`2021`, ~`2022`, ~`2023`, ~`2024`, ~`2025`,
  
  "ATL", "Hartsfield-Jackson Atlanta Intl", "Atlanta",           "United States",  "North America",
  42.92,   75.70,   93.70,  104.65,  108.07,  106.30,
  
  "FRA", "Frankfurt Airport",               "Frankfurt",         "Germany",        "Europe",
  18.77,   24.81,   48.92,   59.36,   61.56,   63.19,
  
  "PKX", "Beijing Daxing Intl",             "Beijing",           "China",          "East Asia",
  16.00,   25.00,   10.23,   39.41,   49.42,   56.00,
  
  "DXB", "Dubai International",             "Dubai",             "UAE",            "Middle East",
  25.84,   29.11,   66.10,   86.90,   92.30,   95.20,
  
  "LHR", "London Heathrow",                 "London",            "United Kingdom", "Europe",
  22.11,   19.39,   61.60,   79.20,   83.86,   84.46,
  
  "DFW", "Dallas/Fort Worth Intl",          "Dallas-Fort Worth", "United States",  "North America",
  39.40,   62.50,   73.40,   81.80,   87.80,   90.00
)

# ------------------------------------------------------------
# 2. Pivot to tidy (long) format
#    One row = one airport × one year
# ------------------------------------------------------------
airports_tidy <- airports_wide |>
  pivot_longer(
    cols      = `2020`:`2025`,
    names_to  = "year",
    values_to = "passengers_millions"
  ) |>
  mutate(year = as.integer(year)) |>
  arrange(iata_code, year)

# ------------------------------------------------------------
# 3. Add year-over-year change columns (within each airport)
# ------------------------------------------------------------
airports_tidy <- airports_tidy |>
  group_by(iata_code) |>
  mutate(
    yoy_change_millions = passengers_millions - lag(passengers_millions),
    yoy_pct_change      = round((passengers_millions / lag(passengers_millions) - 1) * 100, 1)
  ) |>
  ungroup()

# ------------------------------------------------------------
# 4. Enforce factor ordering for airports (by 2024 traffic)
# ------------------------------------------------------------
airport_order <- airports_tidy |>
  filter(year == 2024) |>
  arrange(desc(passengers_millions)) |>
  pull(iata_code)

airports_tidy <- airports_tidy |>
  mutate(iata_code = factor(iata_code, levels = airport_order))

# ------------------------------------------------------------
# 5. Preview
# ------------------------------------------------------------
glimpse(airports_tidy)
print(airports_tidy, n = 36)







# ============================================================
# Busiest Passenger Airports — Line Plot (2020–2025)
# ============================================================

library(tidyverse)

# ------------------------------------------------------------
# Plot-specific wrangling: label data for direct line labels
# (avoids a legend; one label anchored at the 2025 endpoint)
# ------------------------------------------------------------
label_data <- airports_tidy |>
  filter(year == 2025)

# ------------------------------------------------------------
# Plot
# ------------------------------------------------------------
ggplot(airports_tidy, aes(x = year, y = passengers_millions,
                          color = iata_code, group = iata_code)) +
  
  # COVID dip shading
  annotate("rect",
           xmin = 2019.5, xmax = 2021.5,
           ymin = -Inf,   ymax = Inf,
           fill = "grey90", alpha = 0.5) +
  annotate("text",
           x = 2020.5, y = 108,
           label = "COVID-19\nrestrictions",
           size = 3, color = "grey50", fontface = "italic") +
  
  # Lines and points
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  
  # Direct endpoint labels (replaces legend)
  geom_text(data  = label_data,
            aes(label = paste0(iata_code, "\n", passengers_millions, "M")),
            hjust  = -0.1,
            size   = 3.2,
            fontface = "bold",
            show.legend = FALSE) +
  
  # Axis formatting
  scale_x_continuous(
    breaks = 2020:2025,
    expand = expansion(mult = c(0.03, 0.16))   # right margin for labels
  ) +
  scale_y_continuous(
    breaks = seq(0, 120, by = 20),
    labels = scales::label_number(suffix = "M"),
    limits = c(0, 120)
  ) +
  scale_color_brewer(palette = "Dark2") +
  
  # Labels
  labs(
    title    = "Annual Passenger Traffic at Six Major Airports (2020–2025)",
    subtitle = "Total passengers (enplaned + deplaned + direct transit), in millions",
    x        = "Year",
    y        = "Passengers (millions)",
    caption  = "Source: Wikipedia — List of Busiest Airports by Passenger Traffic\n2025 figures are preliminary estimates published by each airport's operator."
  ) +
  
  # Theme
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey40", margin = margin(b = 10)),
    plot.caption     = element_text(color = "grey50", size = 8, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.line.x      = element_line(color = "grey70"),
    legend.position  = "none"
  )


```

## 5.2.2 Monte Carlo Simulation Code

```{r}
monte_carlo_sim <- function(n, x_min, x_max, y_min, y_max, f) {
  x_vals <- runif(n, min = x_min, max = x_max)
  y_vals <- runif(n, min = y_min, max = y_max)
  
  df <- data.frame(
    x = x_vals,
    y = y_vals
  )
  
  df$fx <- f(df$x)
  df$inside <- ifelse(df$y <= df$fx, TRUE, FALSE)
  
  rect_area <- (x_max - x_min) * (y_max - y_min)
  proportion_inside <- mean(df$inside)
  estimate <- proportion_inside * rect_area
  
  attr(df, "estimate") <- estimate
  return(df)
}

library(ggplot2)
library(dplyr)

f <- function(x) dbeta(x, shape1 = 2, shape2 = 5)

df <- monte_carlo_sim(
  n = 1000,
  x_min = 0,
  x_max = 1,
  y_min = 0,
  y_max = 1.5,   
  f = f
)

df <- df %>%
  mutate(category = ifelse(inside, "Below Curve", "Above Curve"))

estimate <- attr(df, "estimate")

curve_data <- data.frame(
  x = seq(0, 1, length.out = 200)
) %>%
  mutate(y = f(x))

plot1 <- ggplot() +
  geom_point(data = df, aes(x = x, y = y, color = category), alpha = 0.6) +
  geom_line(data = curve_data, aes(x = x, y = y), color = "black", size = 1) +
  labs(
    title = paste("Monte Carlo Integration (n = 1000)"),
    subtitle = paste("Estimated Area =", round(estimate, 4)),
    x = "x",
    y = "y",
    color = "Point Type"
  ) +
  theme_minimal()

print(plot1)
```

## 5.2.3 GenAI Original Tidying and Plotting Code

```{r}
# Load required libraries
library(tidyverse)
library(gridExtra)

# ------------------------------------------------------------------------------
# STEP 1: READ THE DATA
# ------------------------------------------------------------------------------

# Read the CSV file
calcium_data <- read.csv('C:/Users/andre/Downloads/calcium.csv', check.names = FALSE)

# View first few rows to understand structure
head(calcium_data)

# ------------------------------------------------------------------------------
# STEP 2: TIDY THE DATA (Convert from wide to long format)
# ------------------------------------------------------------------------------

# The data has 8 columns: 
# First 4 columns (y0, y1, y2, y3) = Group 1
# Last 4 columns (y0, y1, y2, y3) = Group 2

# Extract Group 1 columns (first 4 columns)
group1_data <- calcium_data[, 1:4]
names(group1_data) <- c("y0", "y1", "y2", "y3")

# Extract Group 2 columns (last 4 columns)
group2_data <- calcium_data[, 5:8]
names(group2_data) <- c("y0", "y1", "y2", "y3")

# Add participant ID to each group
group1_data$participant_id <- 1:nrow(group1_data)
group2_data$participant_id <- 1:nrow(group2_data)

# Convert Group 1 to long format
group1_long <- group1_data %>%
  pivot_longer(
    cols = starts_with("y"),
    names_to = "time_point",
    values_to = "measurement"
  ) %>%
  mutate(
    group = "Group A",
    time_numeric = as.numeric(gsub("y", "", time_point))
  ) %>%
  filter(!is.na(measurement))  # Remove any NA values

# Convert Group 2 to long format
group2_long <- group2_data %>%
  pivot_longer(
    cols = starts_with("y"),
    names_to = "time_point",
    values_to = "measurement"
  ) %>%
  mutate(
    group = "Group B",
    time_numeric = as.numeric(gsub("y", "", time_point))
  ) %>%
  filter(!is.na(measurement))  # Remove any NA values

# Combine both groups into one tidy table
tidy_calcium_data <- bind_rows(group1_long, group2_long)

# Display the tidy table structure
print("=== TIDY DATA TABLE ===")
print(head(tidy_calcium_data, 10))
print("\n")
print("Table structure:")
str(tidy_calcium_data)

# ------------------------------------------------------------------------------
# STEP 3: CALCULATE SUMMARY STATISTICS
# ------------------------------------------------------------------------------

# Calculate mean and standard deviation for each group at each time point
summary_stats <- tidy_calcium_data %>%
  group_by(group, time_point, time_numeric) %>%
  summarise(
    mean_value = mean(measurement, na.rm = TRUE),
    sd_value = sd(measurement, na.rm = TRUE),
    n = n(),
    se_value = sd_value / sqrt(n),  # Standard error
    .groups = "drop"
  )

print("\n=== SUMMARY STATISTICS ===")
print(summary_stats)

# ------------------------------------------------------------------------------
# STEP 4: CREATE COMPARISON GRAPHS
# ------------------------------------------------------------------------------

# Graph 1: Boxplot comparison at each time point
graph_boxplot <- ggplot(tidy_calcium_data, aes(x = time_point, y = measurement, fill = group)) +
  geom_boxplot(alpha = 0.7, outlier.size = 1.5) +
  stat_summary(fun = mean, geom = "point", position = position_dodge(width = 0.75), 
               shape = 23, size = 3, fill = "white") +
  labs(
    title = "Comparison of Calcium Measurements: Group A vs Group B",
    subtitle = "Boxplots show distribution, white diamonds show means",
    x = "Time Point",
    y = "Measurement Value",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral"))

# Graph 2: Line plot showing individual trajectories + group mean
graph_individual <- ggplot(tidy_calcium_data, aes(x = time_numeric, y = measurement, group = participant_id)) +
  geom_line(aes(color = group), alpha = 0.3, size = 0.5) +
  geom_point(aes(color = group), alpha = 0.3, size = 1) +
  stat_summary(aes(color = group), fun = mean, geom = "line", size = 1.5) +
  stat_summary(aes(color = group), fun = mean, geom = "point", size = 3) +
  labs(
    title = "Individual Trajectories with Group Means",
    subtitle = "Faint lines = individual participants, Thick lines = group averages",
    x = "Time Point",
    y = "Measurement Value",
    color = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_x_continuous(breaks = 0:3, labels = c("y0", "y1", "y2", "y3")) +
  scale_color_manual(values = c("Group A" = "blue", "Group B" = "red"))

# Graph 3: Bar chart with error bars (mean ± SE)
graph_bars <- ggplot(summary_stats, aes(x = time_point, y = mean_value, fill = group)) +
  geom_bar(stat = "identity", position = position_dodge(0.9), alpha = 0.7) +
  geom_errorbar(aes(ymin = mean_value - se_value, ymax = mean_value + se_value),
                position = position_dodge(0.9), width = 0.25) +
  labs(
    title = "Mean Calcium Measurements (± Standard Error)",
    subtitle = "Bars show means, error bars show ±1 SE",
    x = "Time Point",
    y = "Mean Measurement Value",
    fill = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral"))

# Graph 4: Violin plot with jittered points (shows distribution well)
graph_violin <- ggplot(tidy_calcium_data, aes(x = time_point, y = measurement, fill = group)) +
  geom_violin(alpha = 0.5, position = position_dodge(0.9)) +
  geom_jitter(aes(color = group), position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.9),
              alpha = 0.4, size = 1) +
  stat_summary(fun = mean, geom = "point", position = position_dodge(0.9),
               shape = 18, size = 4, color = "black") +
  labs(
    title = "Distribution of Calcium Measurements",
    subtitle = "Violin plots show density, points show individual values, black diamonds show means",
    x = "Time Point",
    y = "Measurement Value",
    fill = "Group",
    color = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(values = c("Group A" = "lightblue", "Group B" = "lightcoral")) +
  scale_color_manual(values = c("Group A" = "blue", "Group B" = "red"))

# ------------------------------------------------------------------------------
# STEP 5: DISPLAY ALL GRAPHS
# ------------------------------------------------------------------------------

# Print each graph individually
print(graph_boxplot)
print(graph_individual)
print(graph_bars)
print(graph_violin)

# ------------------------------------------------------------------------------
# STEP 6: STATISTICAL COMPARISON
# ------------------------------------------------------------------------------

# Perform t-tests comparing groups at each time point
stat_comparisons <- tidy_calcium_data %>%
  group_by(time_point) %>%
  summarise(
    group_A_mean = mean(measurement[group == "Group A"]),
    group_B_mean = mean(measurement[group == "Group B"]),
    difference = group_A_mean - group_B_mean,
    t_statistic = t.test(measurement ~ group)$statistic,
    p_value = t.test(measurement ~ group)$p.value,
    significance = ifelse(p_value < 0.05, "Significant", "Not significant")
  )

print("\n=== STATISTICAL COMPARISONS (t-tests) ===")
print(stat_comparisons)
```
